---
title: "Continuous Case"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Case}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For this example we are going to walk through how to use Bayesian dynamic borrowing with propensity score matching. The first thing we need is some data. I am going to start with external control data that has come covariates


```{r setup}
library(beastt)
library(tibble)
library(distributional)
library(ggdist)
library(ggplot2)

set.seed(123)

## Make the external dataset
n_external <- 150
sd_external_control = 5

external_df <- tibble(subjid = 1:n_external,   # participant ID
                 cov1 = round(rnorm(n_external, 50, 10)),   # covariate 1
                 cov2 = rbinom(n_external, 1, .2),          # covariate 2  (binary): ~20% level 1
                 cov3 = rbinom(n_external, 1, .6),          # covariate 3  (binary): ~60% level 1
                 cov4 = rbinom(n_external, 1, .3),          # covariate 4  (binary): ~30% level 1
                 y = rnorm(n_external, mean = 45, sd = sd_external_control))
                 

## Make the internal dataset

# Sample sizes
n_inter_control <- 60
n_inter_treat <- 60
n_internal <- n_inter_control + n_inter_treat     # total internal sample size

internal_df <- tibble(subjid = 1:n_internal,                  # participant ID
                  cov1 = round(rnorm(n_internal, 55, 8)),          # covariate 1: mean=55, SD=8 (not mean=50, SD=10)
                  cov2 = rbinom(n_internal, 1, .3),                # covariate 2: ~30% level 1 (not 20%)
                  cov3 = rbinom(n_internal, 1, .5),                # covariate 3: ~50% level 1 (not 60%)
                  cov4 = rbinom(n_internal, 1, .3),                # covariate 4: ~30% level 1 (same)
                  trt = rep(0:1, c(n_inter_control, n_inter_treat)),
                  y = rnorm(n_internal, mean = 50, sd = 3))    # treatment (0: control, 1: active treatment)
```


Now we have some data we can calculate the propensity scores for the internal and external controls using the `calc_prop_scr` function. This creates a propensity score object which we can use for calculating power priors in the next step. 
```{r}
ps_model <- ~ cov1 + cov2 + cov3 + cov4
ps_obj <- calc_prop_scr(internal_df = internal_df, 
                        external_df = external_df, 
                        id_col = subjid,
                          model = ps_model)
prop_scr_hist(ps_obj)
prop_scr_dens(ps_obj, "ipw")

```


We the propensity score object we are able to use that to calculate a power prior. To calculate the power prior, 
```{r}
pwr_prior <- calc_power_prior(prior = dist_normal(92, 100), 
                 ps_obj, response = y, 
                 external_control_sd = 15)

ggplot(tibble(), aes(xdist = pwr_prior, y =1)) +
  stat_slabinterval() + 
  theme_bw()
```

