---
title: "Package Introduction"
package: beastt
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Package Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(beastt)
```

In this vignette, you will learn about the scope of `beastt()` and where to find additional information on how to implement your analysis in `beastt()`.

## Introduction

Bayesian dynamic borrowing (BDB) is a way to incorporate external data into the analysis of data from an internal trial to effectively increase the sample size and in turn increase power. BEASTT focuses on implementing BDB, specifically through the use of inverse probability weighted power priors that can be incorporated into a robust mixtures prior.

`beastt` has the following goals:

- goal 1
- goal 2
- goal 3


## Common Usage

### Create Data Sets

```{r eval=FALSE}
library(beastt)
library(ggdist)
library(ggplot2)

set.seed(1234)

# Create internal and external datasets
internal_df <- data.frame(id_col = 1:20, cov1 = rnorm(10, 2), cov2 = rnorm(100, 20), 
                          y = rnorm(20, mean = 5, sd = 3)
external_df <- data.frame(id_col = 21:40, cov1 = rnorm(10, 2), cov2 = rnorm(100, 18), 
                          y = rnorm(20, mean = 8, sd = 4)

```


### Propensity Score


### Power Prior
