---
title: "Weibull Endpoint with Borrowing from an External Control Arm"
package: beastt
output: 
  rmarkdown::html_vignette:
    toc: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Weibull Endpoint with Borrowing from an External Control Arm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(beastt)
```

In this example, we are going to walk through how to use Bayesian dynamic borrowing (BDB) with the inclusion of inverse probability weighting to balance baseline covariate distributions between external and internal datasets. This particular example considers a hypothetical trial with a cross sectional normal endpoint and a known standard deviation (SD) in each treatment arm (external control arm and both internal arms), and our objective is to use BDB with IPWs to construct a posterior distribution for the control mean $\theta_C$. We will use simulated internal and external datasets from the package where each dataset has a normally distributed response variable and four baseline covariates which we will balance.


```{r class.source = 'fold-hide'}
library(tibble)
library(distributional)
library(dplyr)
library(ggplot2)
set.seed(1234)

summary(int_tte_df)
summary(ex_tte_df)
n_external <- nrow(ex_tte_df)
```

With the covariate data from both the external and internal datasets, we can calculate the propensity scores and ATT inverse probability weights (IPWs) for the internal and external controls using the `calc_prop_scr` function. This creates a propensity score object which we can use for calculating an inverse probability weighted power prior in the next step.

**Note: when reading external and internal datasets into `calc_prop_scr`, be sure to include only the treatment arms across which you want to balance the covariate distributions.** In this example, we want to balance the covariate distributions of the external control arm to be similar to those of the internal control arm, so we will exclude the internal active treatment arm data from this function.

```{r}
ps_model <- ~ cov1 + cov2 + cov3 + cov4
ps_obj <- calc_prop_scr(internal_df = filter(int_tte_df, trt == 0), 
                        external_df = ex_tte_df, 
                        id_col = subjid,
                        model = ps_model)

ps_obj


tidy(ps_obj)
```

In order to check the suitability of the external data, we can create a variety of diagnostic plots. The first plot we might want is a histogram of the overlapping propensity score distributions from both datasets. To get this, we use the `prop_scr_hist` function. This function takes in the propensity score object made in the previous step, and we can optionally supply the variable we want to look at (either the propensity score or the IPW). By default, it will plot the propensity scores. Additionally, we can look at the densities rather than histograms by using the `prop_scr_dens` function. When looking at the IPWs with either the histogram or the density functions, it is important to note that only the IPWs for external control participants will be shown because the ATT IPWs of all internal control participants are equal to 1.

```{r}
prop_scr_hist(ps_obj)
prop_scr_dens(ps_obj, variable = "ipw")
```

The final plot we might want to look at is a love plot to visualize the absolute standard mean differences (both unadjusted and adjusted by the IPWs) of the covariates between the internal and external data. To do this, we use the `prop_scr_love` function. Like the previous function, the only required parameter for this function is the propensity score object, but we can also provide a location along the x-axis for a vertical reference line.

```{r}
prop_scr_love(ps_obj, reference_line = 0.1)
```




```{r}
pwr_prior <- calc_power_prior_weibull(
  ps_obj,
  y, event, 
  dist_normal(0, 10),
  50,
  approximation = "MCMC"
)
plot_dist(pwr_prior)
```

Now that we have a normal power prior, we can calculate the posterior distribution for $\theta_C$ using the `calc_post_norm` function. By defining our prior to be a normal distribution and by assuming the SD of the internal response data to be known, the resulting posterior distribution will also be normal (the case when the SD is unknown is not covered in this vignette).

**Note: if reading internal data directly into `calc_post_norm` instead of a propensity score object, be sure to include only the treatment arm of interest (e.g., the internal control arm if creating a posterior distribution for $\theta_C$).**

```{r}
post <- calc_post_norm(ps_obj,
                       response = y, 
                       prior = pwr_prior,
                       internal_sd = sd_internal_control)
plot_dist(post)
```

If we want to robustify our power prior for $\theta_C$, we can add a vague component to the power prior distribution we previously created to construct a robust mixture prior which we can then pass to `calc_post_norm`. In general, we can define our prior to be a mixture distribution with an arbitrary number of normal and/or $t$ components. If any component of the prior is a $t$ distribution, the prior will be approximated with the mixture of two normal distributions.

```{r}
mixed_prior <- robustify_norm(pwr_prior, n_external)
plot_dist("Robust Mixture Prior" = mixed_prior,
          "Power Prior" = pwr_prior)
```



```{r}
cont_post <- calc_post_weibull(filter(int_tte_df, trt == 0),
                  y, event, rpp, c(10, 12))

vague_only <- dist_mixture(parameters(rpp)$dist, weights = c(0,1))

trt_post <- calc_post_weibull(filter(int_tte_df, trt == 1),
                  y, event, vague_only, c(10, 12))
```


```{r}
ctl_post_ap2 <- as.data.frame(extract(cont_post, pars = c("survProb")))[,1]
trt_post_ap2 <- as.data.frame(extract(trt_post, pars = c("survProb")))[,1]
# Distribution of differences between treatment and control survival probabilities at prespecified time
diff_post_ap2 <- trt_post_ap2 - ctl_post_ap2
mean(diff_post_ap2 > gamma0)    # post prob that diff in surv probs is greater than gamma0
```


